%%%%%%%% ICML 2026 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs} % for professional tables
\usepackage{tikz}
\usepackage{svg}
\usepackage{algorithm2e}



% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2026} with \usepackage[nohyperref]{icml2026} above.
\usepackage{hyperref}


% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
\usepackage{icml2026}

% For preprint, use
% \usepackage[preprint]{icml2026}

% If accepted, instead use the following line for the camera-ready submission:
% \usepackage[accepted]{icml2026}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}


% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Limits of RL for decision tree policies}

\begin{document}

\twocolumn[
  \icmltitle{Limits of reinforcement learning for decision trees in\\ Markov decision processes}

  % It is OKAY to include author information, even for blind submissions: the
  % style file will automatically remove it for you unless you've provided
  % the [accepted] option to the icml2026 package.

  % List of affiliations: The first argument should be a (short) identifier you
  % will use later to specify author affiliations Academic affiliations
  % should list Department, University, City, Region, Country Industry
  % affiliations should list Company, City, Region, Country

  % You can specify symbols, otherwise they are numbered in order. Ideally, you
  % should not use this facility. Affiliations will be numbered in order of
  % appearance and this is the preferred way.
  \icmlsetsymbol{equal}{*}

  \begin{icmlauthorlist}
    \icmlauthor{Hector Kohler}{yyy, sch}
    \icmlauthor{Riad Akrour}{comp, sch}
    \icmlauthor{Philippe Preux}{comp, sch}
    %\icmlauthor{}{sch}
    %\icmlauthor{}{sch}
  \end{icmlauthorlist}

  \icmlaffiliation{yyy}{Laboratiore d'Informatique, \'Ecole Polytechnique, Paris, France}
  \icmlaffiliation{comp}{CRIStAL, Universit\'e de Lille, Lille, France}
  \icmlaffiliation{sch}{Inria, France}

  \icmlcorrespondingauthor{Hector Kohler}{hector.kohler@polytechnique.edu}

  % You may provide any keywords that you find helpful for describing your
  % paper; these are used to populate the "keywords" metadata in the PDF but
  % will not be shown in the document
  \icmlkeywords{Reinforcement learning, decision trees, POMDP}

  \vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column listing the
% affiliations and the copyright notice. The command takes one argument, which
% is text to display at the start of the footnote. The \icmlEqualContribution
% command is standard text for equal contribution. Remove it (just {}) if you
% do not need this facility.

% Use ONE of the following lines. DO NOT remove the command.
% If you have no special notice, KEEP empty braces:
\printAffiliationsAndNotice{}  % no special notice (required even if empty)
% Or, if applicable, use the standard equal contribution text:
% \printAffiliationsAndNotice{\icmlEqualContribution}

\begin{abstract}
  For applications like medicine, machine learning models ought to be interpretable. In that case, models like decision trees are preferred over neural networks because humans can read their predictions from the root to the leaves.
  Learning such decision trees for sequential decision making problems is a relatively new research direction and most of the existing literature focuses on imitating (or distilling) neural networks.
  In contrast, we study reinforcement learning (RL) algorithms that \textit{directly} return decision trees optimizing some trade-off of cumulative rewards and interpretability in a Markov decision process (MDP). 
  We show that such algorithms can be seen as learning policies for partially observable Markov decision processes (POMDPs).
  We use this parallel to understand why in practice it is often easier to use imitation learning than to learn the decision tree from scratch for MDPs.
\end{abstract}

\input{intro}
\input{related}
\input{technicals}
\input{pomdp}
\input{rl_pomdp}

\section{Discussion}
In this paper, we were interested in algorithms that can learn decision tree policies that directly optimize some trade-off of interpretability and performance in MDPs without relying on an oracle or expert policy.
Starting from the IBMDP formulation from~\citet{topin2021iterative}, we have shown that direct learning of decision tree policies for MDPs can be framed as learning deterministic memoryless policies in POMDPs that we called POIBMDPs.
By bridging the gap with the POMDP literature, we conjectured that partial observability is the main limitation for reinforcement learning of decision tree policies for MDPs.
We then supported this conjecture by benchmarking different reinforcement learning algorithms on carefully crafted problems for which we knew the exact optimal decision tree policies.
Across our experiments, we found that only when partial observablity is absent from the learning task can good decision tree policies be trained.
Attempting to overcome the partial observability challenges highlighted so far seems like a bad research direction.
Indeed, while algorithms tailored specifically for the problem of learning deterministic memoryless policies for POIBMDPs might exist, imitation learning works well in practice and has been the state-of-the-art for interpretable sequential decision making for a while.
Furthermore there are other limitations that we did not cover in the framework of~\citet{topin2021iterative} such as how to choose good candidates information-gathering actions or simply how to choose $\zeta$ for a target interpretability-performance trade-off.
Finally, while we focused on non-parametric tree learning assuming RL algorithms should naturally trade off interpretability and performance through the reward signal $\zeta$ for adding nodes to the decision tree policy, another future research avenue could be to develop better algortihms training parametric trees.
Indeed parametric tree policies, on the other hand, can be computed with reinforcement learning directly in the downstream MDP.
However such RL algorithms for parametric decision tree policies~\cite{silva,vos2024optimizinginterpretabledecisiontree,sympol} require to re-train a policy entirely for each desired level of interpretability, i.e. each unique tree structure.
Future research in this direction should focus on algorithms for parametric tree policies that can re-use samples from one tree learning to train a different tree structure more efficiently.
This would reduce the required quantity of a priori knowledge on the decision tree policy structure.
\section*{Impact statement}
Our work put great emphasis on covering existing work, open sourcing code, and being transparent about limitations.
We hope that the impact of our work is going to be positive for society through advancing reseach in interpretable machine learning.
\bibliography{example_paper}
\bibliographystyle{icml2026}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn
\input{appendix}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
% by Iain Murray in 2018, and modified by Alexandre Bouchard in
% 2019 and 2021 and by Csaba Szepesvari, Gang Niu and Sivan Sabato in 2022.
% Modified again in 2023 and 2024 by Sivan Sabato and Jonathan Scarlett.
% Previous contributors include Dan Roy, Lise Getoor and Tobias
% Scheffer, which was slightly modified from the 2010 version by
% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
% slightly modified from Prasad Tadepalli's 2007 version which is a
% lightly changed version of the previous year's version by Andrew
% Moore, which was in turn edited from those of Kristian Kersting and
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.
